Live coding environment for a data exploration language

### 2019-02-02 wip-live-programming/paper.tex
Finish cleaning up intro sections

\begin{figure}[b]
\includegraphics[scale=0.5]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database and adds ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which
make it possible to combine text, equations and code with results of running the code such as tables
or visualizations. Notebooks blur the distinction between development and execution.
Data analysts write small snippets of code, run them to see results immediately and then revise
them.

% interactively based on the results. We aim to build foundations for such seamless integration
% of development and execution in a notebook environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
systems more interactive and spreadsheet-like in such cases is of crucial importance as it enables
more people to produce reproducible and transparent data analyses.

As an example, consider the anlysis done by the Financial Times for an article on plastic
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis.
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

@@

~

TODO: Most logic in external libs

~

Many Jupyter notebooks are more complex than the example discussed here. They might define
helper functions or even structure code using object-oriented programming. However, simple
data analyses such as the one discussed here are frequent enough to pose an interesting and
important domain for programming tools research.

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
    count the number of distinct events per athlete. We then type `.' to choose
    further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment. The table is produced as
      the data anlyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small TBD}
  \label{first-figure-with-subfigures}
\end{figure}

@@

Data exploration of the kind discussed in the previous section has been the motivation
for a recent programming environment called The Gamma. The scripting language in The Gamma
captures the structure discussed above.

A script is a sequence of commands that can either define a
variable or produce an output. The language does not support top-level function declarations,
although we propose an extension that provides a simple abstraction mechanism and is compatible with
providing live previews in Section~X. Lambda functions can be used only as method
arguments.

Given the limited expressiveness of The Gamma, libraries for it are implemented in other languages, such as
JavaScript. More importantly, The Gamma supports type providers, which can be used to generate types,
on the fly, for accessing and exploring external data sources. Type providers provide object types
with members and The Gamma makes using those convenient by providing auto-complete when the user
types the dot symbol (`.') to access a member.

The combination of type providers and auto-complete on~``.'' makes it possible to complete a
large number of data exploration tasks through a very simple interaction of selecting operations from
a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on Olympic medals using a type provider.
The names of identifiers such as \qident{sum Bronze} do not have
any special meaning -- they are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma is more complex than shown here -- it supports methods that can take
functions as arguments -- but the above example illustrates the fact that non-trivial data
exploration can be done using a very simple language.

We complement our theoretical work on foundations of live programming environments for data
exploration with an implementation for The Gamma. As discussed in the previous
section, the assumptions about structure of code that are explicit in The Gamma are also
implicitly present in Python and R data analyses produced by journalists, economists and
other users with other than  programming background. Consequently, our theoretical work
applies to simple data analyses using Python and readers not familiar with The Gamma
can assume that we use a small subset of Python.

@@

% https://www.ft.com/content/360e2524-d71a-11e8-a854-33d6f82e62f8
% https://github.com/ft-interactive/recycling-is-broken-notebooks

\section{Live coding for data exploration}

% TODO: What is live coding? (Show preview while typing!)

This paper presents the foundations of a live coding environment for data exploration that
follows the analysis in the previous section. We consider simple data exploration tasks, such as
those done by journalists and we implement our ideas in the editor for TheGamma.

Figure~\ref{fig:gamma-preview} shows a TheGamma script exporing Olympic medal statistics in
our live coding environment. Before discussing the technical details, it is worth highlighting
a number of important points about our approach.

\begin{itemize}
\item\textbf{Real-world use cases.}
Our work has been motivated by the needs outlined in the previous section. In particular, we
have been focusing on a langauge that is simple, yet sufficient for completing basic data
exploration tasks that a journalist might want to do. We have used our live coding environment
to perform a number of simple data analyses looking at topics such as public government spending,
crises in financial markets and activities of a research institute. Qualitative observations about
our experience are reported in Section~X. Our tools can be used in any modern web browser and are
available as an open-source package.

\item\textbf{Text editor integration.}
Our approach is based on an ordinary text editor that allows unrestricted modification of the
code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
how the user modified the code. This is in contrast with approaches based on structured editing
where the editor tracks what changes to code have been made. Using an ordinary text editor makes
providing live previews harder, but we believe that many users still favor the flexibility of
plain text editors.

\item\textbf{Data stay, code changes.}
Our focus has been on the scenario when input data is a static file that does not change, but the
data analyst modifies the source code. This is in contrast with work on incremental computation,
which focuses on scenarios where code stays the same, but input data change.

That said, many of the methods that we used for evaluating live previews such as maintaining
the dependency graph with nodes for individual syntactic elements are closely related to methods
used in incremental computation research. We do not attempt to combine the two, but note that
this is an interesting future direction.
\end{itemize}

@@

We initially present the data exploration calculus without a static type system. Our main focus
is on the live preview mechanism which can be done equally well without types. We consider
type checking in Section~\ref{sec:types}, because the same mechanism that is used for live previews
can also be used for incremental type-checking.

@@

The use of a dependency graph when evaluating previews makes it possible to cache partial results
through a mechanism discussed in Section~\ref{sec:formal-deps} and Section~\ref{sec:previews-cache}.
The mechanism satisfies two properties. First, if we evaluate a preview using dependency
graph with caching, it is the same as the value we would obtain by evaluating the expression
directly. Second, the evaluation of previews using dependency graphs reuses -- in some cases --
previously evaluated partial results. In other words, we show that the mechanism is correct
and implements an effective optimization. We discuss correctness in this section and evaluate
the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:optimization}.

To show that the evaluated previews are correct, we prove two properties. Correctness
(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a dependency graph is constructed,
when we use it to evaluate previews for a program, the previews are the same as the values we
would obtain by evaluating the program commands directly. Determinacy (Theorem~\ref{thm:determinacy})
guarantees that if we cache a preview for a graph node and update the graph, the preview we would
evaluate using the updated graph would be the same as the cached preview.

@@

\section{Case study}
\label{sec:case}

>>>>

\begin{figure}[b]
\includegraphics[scale=0.5]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database (stored in a local file) and adds ISO country codes (from
  an Excel spreadsheet).}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data analysts increasingly use notebook systems such as Jupyter, which
make it possible to combine text, equations and code with results of running the code such as tables
or visualizations. Notebooks blur the distinction between development and execution.
Data analysts write small snippets of code, run them to see results immediately and then revise
them.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models of physical systems to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
systems more interactive and spreadsheet-like can encourage users to choose programming tools
over spreadsheets and produce reproducible and transparent data analyses.

As an example, the anlysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\vspace{0.5em}
\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides opreations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
  external to the data analysis and may even be implemented in other languages -- many Python
  libraries rely on C++.

@@

\noindent
Many Jupyter notebooks are more complex than the example discussed here. They might define
helper functions or even structure code using object-oriented programming. However, this paper
is motivated by our conviction that simple data analyses such as the one discussed here are frequent
enough to pose an interesting and important domain for programming tools research.

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
      a type provider offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
      the theory presented in this paper. The table is produced as
      the data anlyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
    preview (right).}
  \label{first-figure-with-subfigures}
\end{figure}

@@

Data exploration of the kind discussed in the previous section has been the motivation for a
recent work by \citet{gamma} that develops a small scripting language called The Gamma. The language
follows the structure discussed above. A script is a sequence of commands that can either define a
variable or produce an output. The language does not support top-level function declarations
and lambda functions can be used only as method arguments.

Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma also supports type providers \cite{providers-fsharp},
which can be used to generate types, on the fly, for accessing and exploring external data sources.
Type providers provide object types with members and The Gamma makes using those convenient by
providing auto-complete when the user types the dot symbol (`.') to access a member.

As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
possible to complete a large number of data exploration tasks through a very simple interaction of
selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
any special meaning. They are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about strucutre of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
on a subset of Python.

\subsection{Live coding for data exploration}

The practical implementation work that complements the theory presented in this paper implements a live
coding environment for The Gamma and is discussed in Section~\ref{sec:case} in detail. Earlier work by
\citet{gamma} discussed The Gamma language and the type provider mechanism, but the implementation
provided only a standard text editor with auto-completion. Our work brings the experience of using
The Gamma closer to that of spreadsheets by developing an editor that evaluates scripts
instantaneously in background and displays the results as \emph{live previews}.

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma program consits of a list of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

There is a number of guiding principles that our work follows. First, we allow the analyst
to edit code in an unrestricted form in a text editor. Although structured editors provide an
appearling alternative and make recomputation easier, we prefer the transparency
and openness of plain text. Second, we focus on the scenario when code changes, but input
does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
corner cases that typical analysis involves. In contrast to work on incremental computation,
we do not consider the case when data changes, although supporting interactive data exploration
of streaming data is an interesting future direction.

@@

We initially present the data exploration calculus without a static type system. The live preview
mechanism can be discussed equally well without types. We consider type checking later in
Section~\ref{sec:types} as the mechanism that is used for live previews can also be
used for incremental type-checking.

@@

@@

The use of a dependency graph when evaluating previews makes it possible to cache partial results
through a mechanism discussed in Section~\ref{sec:formal-deps} and Section~\ref{sec:previews-cache}.
The mechanism satisfies two properties. First, if we evaluate a preview using dependency
graph with caching, it is the same as the value we would obtain by evaluating the expression
directly. Second, the evaluation of previews using dependency graphs reuses -- in some cases --
previously evaluated partial results. In other words, we show that the mechanism is correct
and implements an effective optimization. We discuss correctness in this section and evaluate
the effectiveness of the optimization, both theoretically and empirically, in Section~\ref{sec:evaluation}.

To show that the previews are correct, we prove two properties. Correctness
(Theorem~\ref{thm:correcntess}) guarantees that, no matter how a dependency graph is constructed,
when we use it to evaluate previews for a program, the previews are the same as the values we
would obtain by evaluating the program commands directly. Determinacy (Theorem~\ref{thm:determinacy})
guarantees that if we cache a preview for a graph node and update the graph, the preview we would
evaluate using the updated graph would be the same as the cached preview.

@@

\section{Case study / implementation}
\label{sec:case}
